Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-03 article/article.tex
Commit 2024-09-03 22:14

The shortwave solar radiation, reaching Earth's surface, is the main
energy source for the atmosphere and the biosphere, and drives and
governs the climate \citep{Gray2010}. It has direct practical
application in industries related to energy and agricultural production.
The variability of its intensity can impose difficulties in predicting
the yield and in designing the specifications of the appropriate
equipment. Significant portion of research has been focused on
predicting the renewable energy production in a fine timescale and in
near real-time \citep[for a review see][]{Inman2013, Graabak2016}.

@@

Some of the proposed underling mechanisms of those enhancements, have
been summarized by \citet{Gueymard2017}; the most important being the
scattering of radiation on the edges of cumulus clouds. It has been also
suggested that enhancement of GHI can be produced by thin cirrus clouds
though refraction and scattering \citep{Thuillier2013}. Further
investigation with radiative transfer modeling and observations pointed
as the prevailing mechanism, the strong forward Mie scattering through
clouds of low optical depth
\citep{Pecenak2016, Thuillier2013, Yordanov2013, Yordanov2015}. Overall,
enhancement events depend on different interactive factors, which
include cloud thickness, structure and type, and the relative position
of the sun and the clouds \citep{Gueymard2017, Veerman2022}.

On multiple sites, cloud enhancements have been reported that exceed
momentarily the solar constant, resulting in clearness indices above
unit. A summary of extreme cloud enhancement (ECE) cases has been
compiled by \citet{Martins2022}. Cloud enhancements can have also some
practical implications. The intensity and duration of enhancements can
affect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, while ECEs have the potential to
compromise the integrity of photovoltaic plants infrastructure
\citep{DoNascimento2019}. It has also been demonstrated that these
events can interfere in the comparison of ground-based and satellite
observations of radiation \citep{Damiani2018}. Global warming has likely
affected cloud coverage in the last few decades. \citet{Liu2023}
reported increases in cloud cover over the tropical and subtropical
oceans and decreases over most continents, while \citet{Dong2023}
reported decreases over North America and Europe. To our knowledge,
there is no evidence on whether this trend has affected also the number
and strength of CE events.

@@

In this study, we evaluate the effects of CE on GHI by investigating
their frequency of occurrence, intensity, and duration in a thirty-year
period of GHI observations at Thessaloniki, Greece. We used modeled
clear sky irradiance as a baseline to identify cloud enhancements, and
we determined long-term trends of the above-mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

@@

The data used in this study were recorded at the monitoring site of the
Laboratory of Atmospheric Physics, Aristotle University of Thessaloniki,
in Thessaloniki, Greece (\(40^\circ\,38'\,\)N, \(22^\circ\,57'\,\)E,
\(80\,\)m~a.s.l.). The GHI data were measured with a Kipp~\& Zonen CM-21
pyranometer and cover the period 01~January 1994 to 31~December 2023.
During the study period, the pyranometer was independently calibrated
three times at the Meteorologisches Observatorium Lindenberg, DWD,
verifying that the stability of the instrument's sensitivity was better
than \(0.7\,\%\) relative to the initial calibration by the
manufacturer. For the acquisition of radiometric data, the signal of the
pyranometer was sampled at a rate of \(1\,\text{Hz}\) with the mean and
standard deviation of these samples calculated and recorded every
minute. The measurements were corrected for the zero offset (``dark
signal'' in volts), which was calculated by averaging all measurements
recorded for a period of \(3\,\text{h}\), before (morning) or after
(evening) the Sun reaches an elevation angle of \(-10^\circ\). The
signal was converted to irradiance using the ramped value of the
instrument's sensitivity between subsequent calibrations.

To further improve the quality of the irradiance data, a manual
screening was performed, in order to remove inconsistent and erroneous
recordings that can occur stochastically or systematically during the
long operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\). Finally, images from a sky camera
have been used in the manual inspection of the CE identification. The
sky camera operates since 2012 and stores images in 5 min time steps. An
overview of the data analyzed for the study of the CE and ECE events is
given in Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily mean of
the measured GHI.

@@

In this study, we define an event as CE when the measured GHI at ground
level, exceeds the expected value under clear-sky conditions. Similarly,
we define as extreme cloud enhancement events (ECE), the cases when GHI
at ground level exceeds the Total Solar Irradiance (TSI) at the top of
the atmosphere (TOA) for the same SZA. Although the duration of these
bursts can vary from seconds to several minutes, here we are constrained
by the temporal resolution of our data to identify events with duration
of at minimum one-minute.

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OI). The OI is defined as the irradiance
difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we used a simple approach for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}. We created a threshold by using an
appropriate relative and/or an additional constant offset. The initial
results showed that we can detect a big portion of the actual CE events.
However, by inspecting the daily plots of irradiance it became evident
that changes in atmospheric conditions introduced numerous false
positive or false negative results. The main reason for these
discrepancies is the variability of the effects of aerosols and water
vapor which were not taken into account in the two simple methods. To
produce a more representative reference we included the effects of these
factors using a radiative transfer model (RTM). The applied methodology
is discussed in section\nobreakspace{}\ref{rtmcs}.

@@

We approximated the expected clear sky \(\text{GHI}_\text{ref}\) with
the radiative transfer model uvspec, part of libRadtran
\citep{Emde2016}, similarly to the approach used by
\citet{Vamvakas2020}. In uvspec we used the solar spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), the
radiative transfer solver ``disort'' in ``pseudospherical'' geometry and
the ``LOWTRAN'' gas parameterization. The model was run for a range of
variables in order to create a look-up table (LUT) for the estimation of
the cloud-free reference irradiance for each individual observation of
our dataset. In this context, the model was run for SZAs in the range
\(10\) -- \(90^\circ\) with a step of \(0.2^\circ\) and for the
atmospheric profiles of the Air Force Geophysics Laboratory
\citep{Anderson1986} midlatitude summer and midlatitude winter,
representative of the warm and cold seasons.

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements are available since 2003 from
the Cimel sun photometer that belongs to the Aerosol Robotic Network
(AERONET) \citep{Giles2019, Buis1998}. From the observations in the
period 2003 -- 2023 we calculated the monthly climatological means and
standard deviation (\(\sigma\)) for the aerosol optical depth (AOD) at
\(500\,\text{nm}\) and the equivalent height of the water column (WC).
The monthly climatological values of AOD and WC, as well as combinations
with additional offsets of \(\pm1\sigma\) and \(\pm2\sigma\), were used
as inputs to the RTM in the construction of the LUT.

@@

The clear-sky reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using simulations with the RTM based on the long-term
variations of the AOD. As AERONET data start only in 2003, we used for
the period 1994 -- 2005 estimates of changes in AOD at
\(340\,\text{nm}\) derived from a collocated Brewer spectrophotometer
\citep{Kazadzis2007} to calculate the change in
\(\text{GHI}_\text{ref}\) due to aerosols during this period.

>>>>

The shortwave solar radiation, reaching Earth's surface, is the main
energy source for the atmosphere and the biosphere, and drives and
governs the climate \citep{Gray2010}. It has direct practical
application in industries related to energy and agricultural production.
The variability of its intensity can impose difficulties in predicting
the yield and in designing the specifications of the appropriate
equipment. A significant portion of the relevant research has been
focused on predicting the renewable energy production in a fine
timescale and in near real-time \citep[for a review
see][]{Inman2013, Graabak2016}.

@@

Some of the proposed underling mechanisms of those enhancements have
been summarized by \citet{Gueymard2017}; the most important being the
scattering of radiation on the edges of cumulus clouds. It has been also
suggested that enhancement of GHI can be produced by thin cirrus clouds
though refraction and scattering \citep{Thuillier2013}. Further
investigation with radiative transfer modeling and observations pointed
as the prevailing mechanism, the strong forward Mie scattering through
clouds of low optical depth
\citep{Pecenak2016, Thuillier2013, Yordanov2013, Yordanov2015}. Overall,
the appearance of CE events depend on different interactive factors,
which include cloud thickness, structure and type, and the relative
position of the sun and the clouds \citep{Gueymard2017, Veerman2022}.

On multiple sites, cloud enhancements have been reported that exceed
momentarily the solar constant, resulting in clearness indices above
unit. A summary of extreme cloud enhancement (ECE) cases has been
compiled by \citet{Martins2022}. Cloud enhancements can have also some
practical implications. The intensity and duration of enhancements can
affect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, while ECEs have the potential to
compromise the integrity of photovoltaic plants infrastructure
\citep{DoNascimento2019}. It has also been demonstrated that these
events can interfere in the comparison of ground-based and satellite
observations of radiation \citep{Damiani2018}. Global warming has likely
affected cloud coverage in the last few decades. \citet{Liu2023}
reported increases in cloud cover over the tropical and subtropical
oceans and decreases over most continents, while \citet{Dong2023}
reported decreases over North America and Europe. To our knowledge,
there is no evidence of whether this trend has affected also the number
and strength of CE events.

@@

In this study, we evaluate the effects of CE events on GHI by
investigating their frequency of occurrence, intensity, and duration in
a thirty-year record of GHI observations at Thessaloniki, Greece. We
used modeled clear sky irradiance as a baseline to identify cloud
enhancements, and we determined long-term trends of the above-mentioned
metrics, their climatology and some general characteristics. To our
knowledge there are no other studies that provide trends from such a
long dataset.

@@

The data used in this study were recorded at the monitoring site of the
Laboratory of Atmospheric Physics, Aristotle University of Thessaloniki,
in Thessaloniki, Greece (\(40^\circ\,38'\,\)N, \(22^\circ\,57'\,\)E,
\(80\,\)m~a.s.l.). The GHI data were measured with a Kipp~\& Zonen CM-21
pyranometer and cover the period 01~January 1994 to 31~December 2023.
During the study period, the pyranometer was independently calibrated
three times at the Meteorologisches Observatorium Lindenberg, DWD,
verifying that the stability of the instrument's sensitivity was better
than \(0.7\,\%\) relative to the initial calibration by the
manufacturer. For the acquisition of radiometric data, the signal of the
pyranometer was sampled at a rate of \(1\,\text{Hz}\) with the mean and
standard deviation of these samples calculated and recorded every
minute. The measurements were corrected for the zero offset (``dark
signal'' in volts), which was calculated by averaging all measurements
recorded for a period of \(3\,\text{h}\), before (morning) or after
(evening) the Sun reaches an elevation angle of \(-10^\circ\). The
signal was converted to irradiance using the ramped value of the
instrument's sensitivity between subsequent calibrations. We note that
the specific model of the pyranometer is not capable in recording the
instrument's temperature, therefore any temperature changes that may
occur during the day (including those due to sudden increases or
decreases in incidence irradiance) were not considered in the data
reduction procedure. However, the sensitivity of the instrument to
temperature is less than \(\pm1\,\%\) for its operational range
(\(-20^\circ\) -- \(50^\circ\)C).

To further improve the quality of the irradiance data, a manual
screening was performed, to remove inconsistent and erroneous recordings
that can occur stochastically or systematically during the long
operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\). Finally, images from a sky camera
have been used in the manual inspection of the CE identification. The
sky camera operates since 2012 and stores images in 5 min time steps. An
overview of the data analyzed for the study of the CE and ECE events is
given in Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily mean of
the measured GHI.

@@

In this study, we define an event as CE when the measured GHI at ground
level exceeds the expected value under clear-sky conditions. Similarly,
we define as extreme cloud enhancement events (ECE), the cases when GHI
at ground level exceeds the Total Solar Irradiance (TSI) at the top of
the atmosphere (TOA) for the same SZA. Although the duration of these
bursts can vary from seconds to several minutes, here we are constrained
by the temporal resolution of our data to identify events with duration
of at minimum one-minute.

For the detection of CE cases, we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OI). The OI is defined as the irradiance
difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we used a simple approach for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}. We created a threshold by using an
appropriate relative and/or an additional constant offset. The initial
results showed that we can detect a big portion of the actual CE events.
However, by inspecting the daily plots of irradiance it became evident
that changes in atmospheric conditions introduced numerous false
positive or false negative results. The main reason for these
discrepancies is the variability of the effects of aerosols and water
vapor which were not considered in the simple method. To produce a more
representative reference we included the effects of these factors using
a radiative transfer model (RTM). The applied methodology is discussed
in section\nobreakspace{}\ref{rtmcs}.

@@

We approximated the expected clear sky \(\text{GHI}_\text{ref}\) with
the radiative transfer model uvspec, part of libRadtran
\citep{Emde2016}, similarly to the approach used by
\citet{Vamvakas2020}. In uvspec we used the solar spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), the
radiative transfer solver ``disort'' in ``pseudospherical'' geometry and
the ``LOWTRAN'' gas parameterization. The model was run for a range of
variables to create a look-up table (LUT) for the estimation of the
cloud-free reference irradiance for each individual observation of our
dataset. In this context, the model was run for SZAs in the range \(10\)
-- \(90^\circ\) with a step of \(0.2^\circ\) and for the atmospheric
profiles of the Air Force Geophysics Laboratory \citep{Anderson1986}
midlatitude summer and midlatitude winter, representative of the warm
and cold seasons.

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements are available since 2003 from
a Cimel sun photometer which is part of the Aerosol Robotic Network
(AERONET) \citep{Giles2019, Buis1998}. From the observations in the
period 2003 -- 2023 we calculated the monthly climatological means and
standard deviation (\(\sigma\)) for the aerosol optical depth (AOD) at
\(500\,\text{nm}\) and the equivalent height of the water column (WC).
The monthly climatological values of AOD and WC, as well as combinations
with additional offsets of \(\pm1\sigma\) and \(\pm2\sigma\), were used
as inputs to the RTM in the construction of the LUT.

@@

The clear-sky reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using simulations with the RTM to account for the
long-term variations of the AOD. As AERONET data start only in 2003, we
used for the period 1994 -- 2005 estimates of changes in AOD at
\(340\,\text{nm}\) derived from a collocated Brewer spectrophotometer
\citep{Kazadzis2007} to calculate the change in
\(\text{GHI}_\text{ref}\) due to aerosols during this period.
